Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-19 doc/invt.tex
invt.c: add analytical error code
invtpar.h: add invtpar_keymatch
correcting mistakes in invt.tex

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques in free energy simulations.
  %
  They allow one to sample a flat distribution
  along a quantity of interest, $z$,
  by adaptively constructing a bias potential,
  which is the negative potential of mean force (PMF).
  %
  The two algorithms differ by
  the manner of updating the bias potential:
  %
  in the WL case, each simulation step triggers
  an update of the bias potential to only
  the bin containing the current $z$,
  while in the metadynamics case,
  the bias potential of several neighboring bins
  are affected as well.
  %
  The asymptotic error of the resulting PMF
  is determined by the updating magnitude,
  and the rate of the reduction of the magnitude
  over the simulation time.
  %
  In the WL case,
  a simple and effective formula for the magnitude
  is given by the inverse of the number of simulation steps.
  %
  In this study, we shall reexamine the optimality of this prescription,
  and attempt to extend it to the metadynamics case.
  %
  Further, we compared the two algorithms, and found that
  although metadynamics produces a smoother PMF profile,
  it is less effective in reducing errors with the shorter wavelengths.
  %
  Thus, it relies more heavily on the assumption of
  the smoothness of the PMF profile,
  especially with a wide updating window.
\end{abstract}

@@

Free energy calculation\cite{frenkel} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute a distribution, $p^*(z)$,
along a quantity of interest $z$, and to see how
the distribution changes with the external conditions,
such as temperatures and pressures.
%
The negative logarithm $-\log p^*(z)$,
or the potential of mean force (PMF),
defines a free energy.
%
A straightforward approach of computing the PMF
is to run multiple independent simulations
under different conditions,
and then piece together the information
gathered there.
%
This solution is often unsatisfactory
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

A better solution is to artificially alter
the target distribution,
such that we can, in a single simulation, sample
a flat distribution\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002}
along $z$ over a wide range.
%
This alleviates the above problem of local trapping
as the system is now forced to travel along the $z$ direction,
which often represents a slow reaction coordinate.
%
It, however, requires one to efficiently construct a
bias potential for an often unknown system.
%
Note that the bias potential is directly related to the PMF
as the former must precisely offset the latter
to achieve a flat distribution.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{laio2002}
are two widely-used techniques for this purpose,
in which a bias potential is actively build-up
based on the simulation history.
%
In essence, these techniques
encourage the system to leave a region that has been visited
by elevating the bias potential there.
%
To be effective, the updates to bias potential
are made at every few steps,
while a side effect is that the underlying dynamics
is no longer a equilibrium sampling
required by statistical mechanics.
%
Thus, one has to reduce the magnitude of updating
over the course of simulation.

@@

In this study,
we shall give a proof of the inverse-time formula,
and try to generalize it to
updating schemes involving neighboring bins,
such as those used in metadynamics.
%
As a byproduct, we shall show that
the single-bin scheme used by the WL algorithm
is generally one of the best
updating schemes in terms of convergence.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

Generally, we may wish to sample
a nearly flat distribution\cite{dayal2004, trebst2004}, $p_i$,
by adjusting the bias potential $V_i$
on the fly.
%
For example,
we can monitor the histogram accumulated over
a period of time and use it for $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1},
and then update
$V_i$ according to
$$
V_i^{\mathrm{new}}
=
V_i^{\mathrm{old}}
-
\log \frac{ p_i } { \pi_i }.
$$
%
%From Eq. \eqref{eq:pi_p_phi1},
%it is clear that to achieve the target distribution
%$\pi_i = p_i$ requires
%%
%\begin{equation}
%\log p^*_i
%=V_i + \ln p_i + c
%\label{eq:bias_pmf}
%\end{equation}
%%
%%
%In the end of the simulation,
%the PMF, $-\log p^*_i$,
%can be deduced from Eq. \eqref{eq:bias_pmf}.
%
The above inversion process,
adopted by early multicanonical or
entropic sampling\cite{berg1992, lee1993},
can be inconvenient in practice,
for we need to make the period
as long as a multiple of
the autocorrelation time,
which is difficult to estimate for an unknown system.

For later convenience, we introduce
%
\begin{equation}
  v_i \equiv V_i - \ln p^*_i + \ln p_i.
  \label{eq:v_def}
\end{equation}
%
Since the last two terms of Eq. \eqref{eq:v_def}
are constant during the course of simulation,
updates to $V_i$ are equivalent to those to $v_i$.
%
In terms of $v_i$, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should approach a constant of $i$
upon convergence (i.e., when $\pi_i \approx p_i$).

@@

In the WL algorithm\cite{wang2001, wang2001pre},
$v_i$ are updated
in each MC step $t$,
and $v_i(t)$ as a function of $t$
is updated as
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }.
  \label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
We refer to the above scheme as the single-bin scheme,
for it applies only to the bias potential
at the current bin, $i(t)$,
%
By contrast, a multiple-bin scheme
means the general case in which
several neighboring bins are updated.

%The values of $v_j(t)$ at $j \ne i(t)$
%are kept unchanged.

One can show that with a constant $\alpha(t) = \alpha > 0$,
the distribution collected from
the trajectory is identical to $p_i$.
%
However, the continuous update
makes $v_i(t)$ a fluctuating quantity,
and Eq. \eqref{eq:pi_p_v} no longer holds
at all times.
%%
%The source of the deviation is two-fold.
%%
%First, since $v_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to decrease $\alpha$ over time
to reduce the error of $v_i$,
in order to get the unbiased distribution
$\ln p_i^*$ via Eq. \eqref{eq:v_def}.

In the original WL scheme,
the updating magnitude $\alpha$ (or $\ln f$
in the original paper) is kept as a constant
for a period of time,
which is referred to as a stage below.
%
The histogram collected in the stage is monitored.
%
Once the histogram is sufficiently flat,
we are allowed to enter a new stage
with a reduced $\alpha$
(usually half as large as
that in the previous stage).
%
This scheme works well for early stages.
%
However, in later stages, it tends to reduce $\alpha$
too quickly, making the asymptotic error
saturate.

@@

A more effective way
of updating the $\alpha(t)$
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$ is the number of steps
from the beginning of the MC simulation,
which shall be referred to as the ``time'' below.
%
This surprisingly simple formula has attracted
several studies, notably the one by Zhou.
%
There are, however, some reservations about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $C/t$ with a constant $C$
different from $1.0$.

@@

In this section,
we shall derive the optimal $\alpha(t)$
for the single-bin scheme,
Eq. \eqref{eq:wl_update}.
%
To do so,
we shall first express the error of $v(t)$
as a functional of $\alpha(t)$,
and then find the optimal value
by functional variation.

@@

To proceed, we first observe that
the average $\bar v = \sum_{i = 1}^n p_i \, v_i$
increases steadily over time:
%
\begin{equation}
\frac{ d \bar v } { d t }
=
\sum_{i = 1}^n p_i \dot v_i
=
\alpha(t) \sum_{i = 1}^n h_i(t) = \alpha(t).
\label{eq:dvbardt}
\end{equation}
%
However, the difference between $v_i$ and $\bar v$,
%
\begin{equation}
  x_i \equiv v_i - \bar v = v_i - \sum_{j = 1}^n p_j \, v_j,
  \label{eq:x_def}
\end{equation}
%
is expected to be small in the asymptotic regime,
and we can expand Eq. \eqref{eq:h_ave} as
\begin{align}
\langle h_i(t) \rangle
&\approx
\frac{ p_i \, e^{- x_i} }
{ \sum_{ j = 1}^n p_j \, e^{- x_j} }
\approx
\frac{ p_i ( 1 - x_i ) }
{ \sum_{ j = 1}^n p_j (1 - x_j) }
\notag \\
&\approx
p_i \, \left(
  1 - x_i + \sum_{j=1}^n p_j \, x_j
\right)
=
p_i \, (1 - x_i),
\label{eq:hdet_linearize}
\end{align}
where we have used $\sum_{j=1}^n p_j = 1$,
and
%
\begin{equation}
  \sum_{i = 1}^n p_i \, x_i = 0,
  \label{eq:px_sum}
\end{equation}
which follows directly from Eq. \eqref{eq:x_def}.
%
With Eqs.
\eqref{eq:vt_diffeq},
\eqref{eq:h_split},
\eqref{eq:dvbardt},
and
\eqref{eq:hdet_linearize},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \frac{ \zeta_i(t) } { p_i } \right].
  \label{eq:dxdt_WL}
\end{equation}

The total error of the bias potential can be written in
terms of $x_i$,
\begin{equation}
\mathcal E
=
\sum_{i = 1}^n p_i \, \langle x_i^2 \rangle
=
\sum_{i = 1}^n p_i \, \left\langle (v_i - \bar v)^2 \right\rangle,
\label{eq:error_sum}
\end{equation}
where
$\bar v = \sum_{i = 1}^n p_i v_i$.
%
Our aim is to find the $\alpha(t)$
that minimizes Eq. \eqref{eq:error_sum}.
%
To do so, we shall first consider
a simpler one-variable problem.

@@

Consider the following equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \lambda \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is a generalized noise
that is invariant under time translation:
%
\begin{equation}
\left\langle \xi(t) \, \xi(t') \right\rangle
=
\kappa(t - t').
\label{eq:noise_correlation}
\end{equation}
%
Additionally, we require
\begin{equation}
  \lim_{t \rightarrow \pm\infty} \kappa(t) = 0.
  \label{eq:kappat_limit}
\end{equation}
%
For example, for a white noise,
$\kappa(t)$ is proportional to
Dirac's $\delta$-function, $\delta(t)$.
%
We wish to show that the optimal $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{1}{\lambda \, (t + t_0)}.
\label{eq:alpha_opt}
\end{equation}

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(t) = x(0) \, e^{-\lambda \, q(t)}
+ \int_0^t \dot u\bigl( q(t') \bigr) \, \xi(t') \, dt',
\label{eq:xt_solution}
\end{equation}
%
where,
%
\begin{equation}
q(t) \equiv \int_0^t \alpha(t') \, dt',
\label{eq:qt_definition}
\end{equation}
%
and
%
\begin{align}
u(z)
&\equiv
e^{-\lambda \, q(t) + \lambda \, z}.
\label{eq:ut_definition}
\end{align}

We shall further demand that
%
\begin{equation}
  \lim_{t \to \infty} q(t) \to \infty.
  \label{eq:qt_limit}
\end{equation}
%
Then, for a large time $t$,
the first term on the right-hand side
of Eq. \eqref{eq:xt_solution} can be neglected, and
%
\begin{align}
\left\langle x^2(t) \right\rangle
%&=
%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%\notag
%\\
&=
\int_0^t \int_0^t
  \dot u\bigl( q(t') \bigr) \,
  \dot u\bigl( q(t'') \bigr) \,
  \kappa(t' - t'') \, dt'' \, dt'.
\label{eq:x2t_average}
\end{align}

@@

Going back to the problem of
minimizing Eq. \eqref{eq:error_sum}
for the single-bin scheme,
we find from Eq. \eqref{eq:dxdt_WL}
that with $\lambda = 1$,
Eq. \eqref{eq:alpha_opt}
will optimize each term $\langle x_i^2 \rangle$,
and hence the sum.
%
Thus, Eq. \eqref{eq:alpha_invt} is optimal
up to a constant $t_0$. %defining the origin of $t$.

@@

We now consider the multiple-bin scheme.
%
In this case,
upon a visit to bin $j$,
we update not only the bias potential there,
but also those for a few neighboring bins $i$.
%
The update magnitudes are given by a matrix $\mathbf w$
with elements $w_{ij}$.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
\begin{equation}
  \dot v_i(t) =
  \sum_{j=1}^n \alpha(t) \, h_j(t) \frac{ w_{ij} } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

The procedure of optimization is largely
the same as the single-bin case,
whereas the matrix nature brings about
some technical difficulties that require
certain restrictions and approximations.

\subsubsection{Updating matrix}

The matrix $\mathbf w$ is not arbitrary.
%
A necessary condition to sample desired distribution
$\mathbf p = (p_1, \dots, p_n)$
is that in Eq. \eqref{eq:vt_diffeq_mbin}
when $h_j(t)$ coincides with $p_j$,
the rate of change $\dot v_i(t)$
should be independent of $i$.
%
This allows $v_i(t)$ to approach a constant
asymptotically.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1.
  \label{eq:w_sumj}
\end{equation}
%
This means $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
although the matrix elements can be negative.

@@

Further, by Eq. \eqref{eq:w_detailedbalance},
we can define a symmetric matrix $\hat{\mathbf w}$
as $\hat w_{ij} = \sqrt{p_i/p_j} \, w_{ij}$
that can be diagonalized
with a set of orthonormal eigenvectors:
%
$\sum_{i = 1}^n \varphi_{ki} \, \hat w_{ij} = \lambda_k \, \varphi_{kj}$.
%
Thus,
in terms of diagonalizing $\mathbf w$, we have
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij} = \lambda_k \, \phi_{kj},
  \label{eq:eig_w}
\end{equation}
where
$\phi_{ij} = \sqrt{p_j} \, \varphi_{ij}$ satisfies
the orthonormal conditions:
%
\begin{align}
\sum_{k = 1}^n \phi_{ki} \, \phi_{kj}
&= \delta_{ij} \, p_i,
\label{eq:eig_orthonormal_cols}
\\
\sum_{k = 1}^n \frac{ \phi_{ik} \, \phi_{jk} }{ p_k }
&= \delta_{ij}.
\label{eq:eig_orthonormal_rows}
\end{align}

%It is worth pointing out
Note that, in a stable updating scheme,
all eigenvalues $\lambda_k$ must be nonnegative,
for otherwise there is a fluctuating mode
that can increase indefinitely.

>>>>

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques in free energy simulations.
  %
  They allow one to sample a flat distribution
  along a quantity of interest, $z$,
  by adaptively constructing a bias potential
  that offsets the potential of mean force (PMF).
  %
  The two algorithms differ by
  the manner of updating the bias potential:
  %
  in the WL case, each simulation step triggers
  an update of the bias potential to only
  the bin containing the current $z$,
  while in the metadynamics case,
  the bias potential of several neighboring bins
  are affected as well.
  %
  The asymptotic error of the resulting PMF
  is determined by the updating magnitude,
  and the rate of the reduction of the magnitude
  over the simulation time.
  %
  In the WL case,
  a simple and effective formula for the magnitude
  is given by the inverse of the number of simulation steps.
  %
  In this study, we shall reexamine the optimality of this prescription,
  and attempt to extend it to the metadynamics case.
  %
  Further, we compared the two algorithms, and found that
  although metadynamics produces a smoother PMF profile,
  which is necessary for stable numerical differentiation
  in computing the force,
  it is less effective in reducing errors with the shorter wavelengths.
  %
  Thus, it relies more heavily on the assumption of
  the smoothness of the PMF profile,
  especially with a wide updating window.
\end{abstract}

@@

Free energy calculation\cite{frenkel} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute a distribution, $p^*(z)$,
along a quantity of interest $z$, and to see how
the distribution changes with external conditions,
such as temperatures and pressures.
%
The negative logarithm $-\log p^*(z)$,
or the potential of mean force (PMF),
defines a free energy.
%
A straightforward approach of computing the PMF
is to run multiple independent simulations
under different conditions,
and then piece together the information
gathered there.
%
This solution is often unsatisfactory
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

A better solution is to artificially alter
the target distribution,
such that we can, in a single simulation, sample
a flat distribution\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002}
along $z$ over a wide range.
%
This alleviates the above problem of local trapping
as the system is now forced to travel more flexibly
along the $z$ direction,
which often represents a slow reaction coordinate.
%
It, however, requires one to efficiently construct a
bias potential for an often unknown system.
%
Note that the bias potential must exactly cancel the PMF
to achieve a flat distribution.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{laio2002}
are two widely-used techniques for this purpose,
in which a bias potential is actively built-up on-the-fly
during simulation.
%
In essence, these techniques
encourage the system to leave a region that has been visited
by elevating the bias potential there.
%
A main difference of the two techniques
is that in the WL case, an update is limited to the bin
containing the current $z$,
while metadynamics adopts a more widespread window
covering several neighboring bins.
%
The latter ensures a smooth profile
suitable for numerical differentiation for force calculation,
and hence is more often used molecular dynamics simulations.
%
To be effective, the updates to bias potential
are made at every few steps,
while a side effect is that the underlying dynamics
is no longer a equilibrium sampling
required by statistical mechanics.
%
Thus, one has to reduce the magnitude of updating
over the course of simulation.

@@

In this study,
we shall give a proof of the inverse-time formula,
and try to generalize it to
updating schemes involving several neighboring bins,
such as those used in metadynamics.
%
As a byproduct, we shall show that
the single-bin scheme used by the WL algorithm
is generally one of the best
updating schemes in terms of convergence.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

Generally, we may wish to sample
a nearly flat distribution\cite{dayal2004, trebst2004}, $p_i$,
by adjusting the bias potential $V_i$
on the fly.
%
For example,
we can monitor the histogram accumulated over
a period of time and use it for $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1},
and then update
$V_i$ according to
$$
V_i^{\mathrm{new}}
=
V_i^{\mathrm{old}}
-
\log \frac{ p_i } { \pi_i }.
$$
%
%From Eq. \eqref{eq:pi_p_phi1},
%it is clear that to achieve the target distribution
%$\pi_i = p_i$ requires
%%
%\begin{equation}
%\log p^*_i
%=V_i + \ln p_i + c
%\label{eq:bias_pmf}
%\end{equation}
%%
%%
%In the end of the simulation,
%the PMF, $-\log p^*_i$,
%can be deduced from Eq. \eqref{eq:bias_pmf}.
%
The above inversion process,
adopted by early multicanonical or
entropic sampling\cite{berg1992, lee1993},
can be inconvenient in practice.
%
For example, a reasonably precise the histogram
requires an accumulation period
longer than the autocorrelation time,
whereas the latter is difficult to
estimate beforehand for an unknown system.

For later convenience, we introduce a shifted bias potential
%
\begin{equation}
  v_i \equiv V_i - \ln p^*_i + \ln p_i.
  \label{eq:v_def}
\end{equation}
%
Since the last two terms of Eq. \eqref{eq:v_def}
remain constant during the course of simulation,
updates to $V_i$'s are equivalent to those to $v_i$'s.
%
In terms of $v_i$'s, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should approach a constant of $i$
upon convergence (i.e., when $\pi_i \approx p_i$).

@@

In the WL algorithm\cite{wang2001, wang2001pre},
$v_i$ are updated
in each MC step $t$,
and $v_i(t)$ as a function of $t$
is updated as
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }.
  \label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
We refer to the above scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$,
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i },
  \label{eq:mbin_update}
\end{equation}
%
where $w_{ij}$ are a set of coefficients satisfying
$\sum_{j=1}^n w_{ij} = 1$.
%
We shall discuss other restrictions on $w_{ij}$
in Sec. \ref{sec:updating-matrix}.

One can show that with a constant $\alpha(t) = \alpha > 0$,
the distribution collected from
the trajectory is identical to $p_i$.
%
However, the above updates
change $v_i(t)$ continuously,
and Eq. \eqref{eq:pi_p_v} no longer holds
at all times.
%%
%The source of the deviation is two-fold.
%%
%First, since $v_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to decrease $\alpha$ over time
to reduce the error of $v_i$,
in order to get the unbiased distribution
$\ln p_i^*$ via Eq. \eqref{eq:v_def}.

In the original WL scheme,
the updating magnitude $\alpha$ (or $\ln f$
in the original paper) is kept as a constant
for a period of time,
which is referred to as a stage below.
%
The histogram collected in the stage is monitored.
%
Once the histogram is sufficiently flat,
we are allowed to enter a new stage
with a reduced $\alpha$\cite{wang2001, wang2001pre}
(usually half as large as
that in the previous stage).
%
This scheme works well for early stages.
%
However, in later stages, it tends to reduce $\alpha$
too quickly, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.

@@

A more effective way
of updating the $\alpha(t)$
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$ is the number of steps
from the beginning of the MC simulation,
which shall be referred to as the ``time'' below.
%
This surprisingly simple formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $C/t$ with a constant $C$
different from $1.0$\cite{
morozov2007, zhou2008, morozov2009}.

@@

In this section,
we shall derive the optimal $\alpha(t)$
for the single-bin scheme,
Eq. \eqref{eq:wl_update}.
%
To do so,
we shall first express the error of $v(t)$
as a functional of $\alpha(t)$,
and then minimize it by variation.

@@

To proceed, we first observe that
the average $\bar v = \sum_{i = 1}^n p_i \, v_i$
increases steadily over time:
%
\begin{equation}
\frac{ d \bar v } { d t }
=
\sum_{i = 1}^n p_i \dot v_i
=
\alpha(t) \sum_{i = 1}^n h_i(t) = \alpha(t).
\label{eq:dvbardt}
\end{equation}
%
However, we expect the difference between $v_i$ and $\bar v$,
%
\begin{equation}
  x_i \equiv v_i - \bar v = v_i - \sum_{j = 1}^n p_j \, v_j,
  \label{eq:x_def}
\end{equation}
%
to be small in the asymptotic regime,
and expand Eq. \eqref{eq:h_ave} as
\begin{align}
\langle h_i(t) \rangle
&\approx
\frac{ p_i \, e^{- x_i} }
{ \sum_{ j = 1}^n p_j \, e^{- x_j} }
\approx
\frac{ p_i ( 1 - x_i ) }
{ \sum_{ j = 1}^n p_j (1 - x_j) }
\notag \\
&\approx
p_i \, \left(
  1 - x_i + \sum_{j=1}^n p_j \, x_j
\right)
=
p_i \, (1 - x_i),
\label{eq:hdet_linearize}
\end{align}
where we have used $\sum_{j=1}^n p_j = 1$,
and
%
\begin{equation}
  \sum_{i = 1}^n p_i \, x_i = 0,
  \label{eq:px_sum}
\end{equation}
which follows directly from Eq. \eqref{eq:x_def}.
%
From Eqs.
\eqref{eq:vt_diffeq},
\eqref{eq:h_split},
\eqref{eq:dvbardt}-\eqref{eq:hdet_linearize},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \frac{ \zeta_i(t) } { p_i } \right].
  \label{eq:dxdt_WL}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:dvbardt}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left( \frac{ h_i } { p_i } - 1 \right)
  \\
  &\stackrel{\mathrm{Eq.\; \eqref{eq:h_split}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left( \frac{ \langle h_i \rangle } { p_i } + \frac{ \zeta_i } { p_i } - 1 \right)
  \stackrel{\mathrm{Eq.\; \eqref{eq:hdet_linearize}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \frac{ \zeta_i } { p_i } \right].
\end{aligned}
$$
}

The total error of the bias potential can be written in
terms of $x_i$,
\begin{equation}
\mathcal E
=
\sum_{i = 1}^n p_i \, \langle x_i^2 \rangle
=
\sum_{i = 1}^n p_i \, \left\langle (v_i - \bar v)^2 \right\rangle,
\label{eq:error_sum}
\end{equation}
where
$\bar v = \sum_{i = 1}^n p_i v_i$.
%
Our aim is to find the $\alpha(t)$
that minimizes Eq. \eqref{eq:error_sum}.
%
Fortunately, we shall find that it suffices to consider
a simpler one-variable problem
in this case.

@@

Consider the following equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \lambda \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is a generalized noise
that is invariant under time translation:
%
\begin{equation}
\left\langle \xi(t) \, \xi(t') \right\rangle
=
\kappa(t - t').
\label{eq:noise_correlation}
\end{equation}
%
Additionally, we require
\begin{equation}
  \lim_{t \rightarrow \pm\infty} \kappa(t) = 0.
  \label{eq:kappat_limit}
\end{equation}
%
For example, for a white noise,
$\kappa(t)$ is proportional to
Dirac's $\delta$-function, $\delta(t)$.
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{1}{\lambda \, (t + t_0)}.
\label{eq:alpha_opt}
\end{equation}

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(t) = x(0) \, e^{-\lambda \, q(t)}
+ \int_0^t \dot u\bigl( q(t') \bigr) \, \xi(t') \, dt',
\label{eq:xt_solution}
\end{equation}
%
where,
%
\begin{equation}
q(t) \equiv \int_0^t \alpha(t') \, dt',
\label{eq:qt_definition}
\end{equation}
%
and
%
\begin{align}
u(z)
&\equiv
e^{-\lambda \, q(t) + \lambda \, z}.
\label{eq:ut_definition}
\end{align}
So
\begin{align}
\left\langle x^2(t) \right\rangle
&=
\int_0^t \int_0^t
  \dot u\bigl( q(t') \bigr) \,
  \dot u\bigl( q(t'') \bigr) \,
  \kappa(t' - t'') \, dt'' \, dt'
\notag
\\
&\hphantom{=}
+\langle x^2(0) \rangle e^{-2 \, \lambda \, q(t)}.
\label{eq:x2t_average}
\end{align}

We shall further demand that
%
\begin{equation}
  \lim_{t \to \infty} q(t) \to \infty.
  \label{eq:qt_limit}
\end{equation}
%
Then, for a large $t$,
the second term on the right-hand side
of Eq. \eqref{eq:xt_solution} can be neglected, and
%
\begin{align}
\left\langle x^2(t) \right\rangle
%&=
%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%\notag
%\\
&=
\int_0^t \int_0^t
  \dot u\bigl( q(t') \bigr) \,
  \dot u\bigl( q(t'') \bigr) \,
  \kappa(t' - t'') \, dt'' \, dt'.
\label{eq:x2t_average_asym}
\end{align}

@@

Going back to the problem of
minimizing Eq. \eqref{eq:error_sum}
for the single-bin scheme.
%
Note that Eq. \eqref{eq:dxdt_WL}
is of the form of Eq. \eqref{eq:dxdt_alpha}.
%
Thus, with $\lambda = 1$,
Eq. \eqref{eq:alpha_opt}
will optimize each term $\langle x_i^2 \rangle$,
and hence the sum.
%
In other words, the $\alpha(t)$ given by
Eq. \eqref{eq:alpha_invt} is optimal
up to a constant $t_0$.

@@

We now consider the multiple-bin scheme,
Eq. \eqref{eq:mbin_update}.
%
In this case,
upon a visit to bin $j$,
we update not only the bias potential there,
but also those for a few neighboring bins $i$.
%
The update magnitudes are given by a matrix $\mathbf w$
with elements $w_{ij}$.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
\begin{equation}
  \dot v_i(t) =
  \sum_{j=1}^n \alpha(t) \, h_j(t) \frac{ w_{ij} } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

The optimization procedure is similar to the single-bin case,
whereas the matrix nature brings about
some technical complexities that require
certain restrictions and approximations.

\subsubsection{\label{sec:updating-matrix}
Updating matrix}

The matrix $\mathbf w$ is not arbitrary.
%
A necessary condition to sample desired distribution
$\mathbf p = (p_1, \dots, p_n)$
is the following.
%
In Eq. \eqref{eq:vt_diffeq_mbin},
if $h_j(t)$ were the same as $p_j$,
the rates of change $\dot v_i(t)$'s
should be independent of $i$.
%
This allows $v_i(t)$'s to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1.
  \label{eq:w_sumj}
\end{equation}
%
The above condition means that $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
although the matrix elements can be negative.

@@

Further, by Eq. \eqref{eq:w_detailedbalance},
we can define a symmetric matrix $\hat{\mathbf w}$
as $\hat w_{ij} = \sqrt{p_i/p_j} \, w_{ij}$
that can be diagonalized\cite{vankampen}
with a set of orthonormal eigenvectors:
%
$\sum_{i = 1}^n \varphi_{ki} \, \hat w_{ij} = \lambda_k \, \varphi_{kj}$.
%
Thus,
in terms of diagonalizing $\mathbf w$, we have
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij} = \lambda_k \, \phi_{kj},
  \label{eq:eig_w}
\end{equation}
where
$\phi_{ij} = \sqrt{p_j} \, \varphi_{ij}$ satisfies
the orthonormal conditions\cite{vankampen}:
%
\begin{align}
\sum_{k = 1}^n \phi_{ki} \, \phi_{kj}
&= \delta_{ij} \, p_i,
\label{eq:eig_orthonormal_cols}
\\
\sum_{k = 1}^n \frac{ \phi_{ik} \, \phi_{jk} }{ p_k }
&= \delta_{ij}.
\label{eq:eig_orthonormal_rows}
\end{align}

%It is worth pointing out
Note that, in a stable updating scheme,
all eigenvalues $\lambda_k$ of $\mathbf w$
must be nonnegative,
for otherwise there is a fluctuating mode
that can increase indefinitely.
